Convergence rate of MAP estimates \\ for the exponential family

### 2021-11-12 aistats.tex
Use running authors because only first row of authors was appearing
corrected typos + updated affiliation

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

@@

\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
\aistatsaddress{SAIT AI Lab, Montreal 
	\And Mila, Université de Montreal \\
	SAIT AI Lab, Montreal\\
	Canada CIFAR AI Chair
}

@@

\paragraph{Models}
Exponential families are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
They are particularly appreciated for their convex log-likelihood
\alignn{
f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.

@@

In general, if $X$ is part of the exponential family,  $T(X)$ is part of the natural exponential family with the appropriate support and base measure, and with the same log-partition function as $X$, up to constants.
MLE and MAP only depend on $T(X)$, not $X$, so their performance only depends on the distributions of $T(X)$.

@@

To obtain the result for Gaussian variance, theorem~\ref{thm:varianceMLE}, simply plug-in $\alpha=\half$.

In this section, we review useful properties of the gamma distribution and associated Bregman divergence in \S\ref{app:gamma}. 
Then we prove theorem~\ref{thm:gammaMLE} in \S\ref{app:gammaMLE}.
Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}.
Then we prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}, in order to prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
Finally we numerically investigate the effect of prior hyper-parameters in \S\ref{app:prior-choice}.

@@

	\begin{proof}
	Since $T(X)$ follows a gamma distribution $\Gamma(\alpha, \beta)$, the MLE is a scaled sum of gammas $\hat \mu_n = \inv{n} \sum_i T(X_i)$.
	As such it is also a gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
	If we consider the ratio $\frac{\MAPm}{\mu_*}$, it is also a gamma with parameter $\Gamma(n \alpha, n \alpha)$.
	Its inverse follows an inverse gamma distribution with expectation
	\begin{equation}
		\expect{\frac{\mu_*}{\hat \mu_n}} 
		= \begin{cases}
			\frac{n\alpha}{n \alpha -1 }\ \text{ if } n\alpha > 1, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{equation}
	which implies that for $n\alpha>1$, 
	\begin{align}
		\expect{\frac{\mu_*}{\hat \mu_n}} -1  
		= \frac{n \alpha}{n\alpha -1 } - 1
		= \frac{1}{n \alpha -1}
	\end{align}
	There is also a closed form solution for the expected logarithm of a gamma.
	Indeed a the sufficient statistic of a gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a gamma to get
	\begin{align}
		\expect{\log \frac{\hat \mu_n}{\mu_*}} 
		= \psi\paren{n \alpha} - \log\paren{n \alpha},
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	%Closed forms also exist for non-central chi-squares, see for example the note of \citep{pav2015moments}.
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \hat \mu_n) }
	&= \alpha \expect{\frac{\mu_*}{ \hat \mu_n} - 1 + \log \left(\frac{\hat \mu_n}{\mu_*} \right) },
	\\
	& =
	\begin{cases}
		\alpha \paren{ \frac{1}{n\alpha - 1} + \psi\paren{n \alpha} - \log\paren{n \alpha} }, \ \text{ if } n\alpha>1, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	Surprisingly, for a gaussian variance where $\alpha = \half$, we need $3$ samples or more for the expected loss to be bounded.
	When the expectation is finite, we can get a more interpretable formula using known \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function},
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x} = - \inv{x} + \inv{2x}	\; ,
	\end{align}
	giving, for $n \alpha >1 $,
	\begin{align}
		\frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha}
		&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
		\leq \frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha} + \frac{\alpha}{2n\alpha}
		\\
		\iff
			\frac{1}{n(n \alpha-1)}
			&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
			\leq \inv{2n} + \frac{1}{n(n \alpha-1)} \; ,
	\end{align}
	so we get a $\Omega(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	\end{proof}

@@

\begin{theorem}[MAP Bound]
Consider an exponential distribution with sufficient statistics coming from a gamma distribution $\Gamma(\alpha, \beta^*)$ with mean parameter $\mu^* = \frac{\alpha}{\beta^*}$.
The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP  with prior hyper-parameters $(n_0,\mu_0)$ is upper bounded as
\alignn{
    \forall n\geq \inv{\alpha},\quad \quad
	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
	\leq\frac{1}{n_0+n} + 
	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} - 1 }^2 }{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} +  \frac{n - \inv{\alpha}}{n_0}}}
}
\end{theorem} 
Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
Expressions for $n\alpha < 1$ are less elegant as visible below
\alignn{
 	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
	\leq \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}} -1 + \frac{a - n_0}{n_0+n}
	\quadtext{where}
	 a:=n_0\frac{\mu_0}{\mu^*} \; .
}

>>>>

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
\runningauthor{R\'emi Le Priol, Frederik Kunstner, Damien Scieur, and Simon Lacoste-Julien}

@@

\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
\aistatsaddress{Samsung, SAIT AI Lab, Montreal 
	\And Mila, Université de Montreal \\
	Samsung, SAIT AI Lab, Montreal\\
	Canada CIFAR AI Chair
}

@@

\paragraph{Models}
Exponential families are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
They are particularly appreciated for their convex log-likelihood
\alignn{
f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is the convex log-partition function and \mbox{$\nat\in\Theta$} is the \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.

@@

In general, if $X$ is part of the exponential family, then $T(X)$ is part of the natural exponential family with the appropriate support and base measure, with the same log-partition function as $X$ up to constants.
MLE and MAP only depend on $T(X)$, not $X$, so their performance only depends on the distribution of $T(X)$.

@@

To obtain the result for Gaussian variance (see Theorem~\ref{thm:varianceMLE}), it suffices to set $\alpha=\half$ in Theorem~\ref{thm:gammaMLE}.

In this section, we review useful properties of the gamma distribution and associated Bregman divergence in \S\ref{app:gamma}. 
Then we prove theorem~\ref{thm:gammaMLE} in \S\ref{app:gammaMLE}.
Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}, and prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}, in order to prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
Finally we numerically investigate the effect of prior hyper-parameters in \S\ref{app:prior-choice}.

@@

	\begin{proof}
	Since $T(X)$ follows a gamma distribution $\Gamma(\alpha, \beta)$, the MLE is a scaled sum of gammas $\hat \mu_n = \inv{n} \sum_i T(X_i)$.
	As such it is also a gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
	If we consider the ratio $\frac{\MAPm}{\mu_*}$, it is also a gamma with parameter $\Gamma(n \alpha, n \alpha)$.
	Its inverse follows an inverse gamma distribution with expectation
	\begin{equation}
		\expect{\frac{\mu_*}{\hat \mu_n}} 
		= \begin{cases}
			\frac{n\alpha}{n \alpha -1 }\ \text{ if } n\alpha > 1, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{equation}
	which implies that for $n\alpha>1$, 
	\begin{align}
		\expect{\frac{\mu_*}{\hat \mu_n}} -1  
		= \frac{n \alpha}{n\alpha -1 } - 1
		= \frac{1}{n \alpha -1}
	\end{align}
	There is also a closed form solution for the expected logarithm of a gamma.
	Indeed, the sufficient statistic of a gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a gamma to get
	\begin{align}
		\expect{\log \frac{\hat \mu_n}{\mu_*}} 
		= \psi\paren{n \alpha} - \log\paren{n \alpha},
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	%Closed forms also exist for non-central chi-squares, see for example the note of \citep{pav2015moments}.
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \hat \mu_n) }
	&= \alpha \expect{\frac{\mu_*}{ \hat \mu_n} - 1 + \log \left(\frac{\hat \mu_n}{\mu_*} \right) },
	\\
	& =
	\begin{cases}
		\alpha \paren{ \frac{1}{n\alpha - 1} + \psi\paren{n \alpha} - \log\paren{n \alpha} }, \ \text{ if } n\alpha>1, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	Surprisingly, for a gaussian variance where $\alpha = \half$, we need $3$ samples or more for the expected loss to be bounded.
	When the expectation is finite, we can get a more interpretable formula using known \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function},
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x} = - \inv{x} + \inv{2x}	\; ,
	\end{align}
	giving, for $n \alpha >1 $,
	\begin{align}
		\frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha}
		&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
		\leq \frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha} + \frac{\alpha}{2n\alpha}
		\\
		\iff
			\frac{1}{n(n \alpha-1)}
			&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
			\leq \inv{2n} + \frac{1}{n(n \alpha-1)} \; ,
	\end{align}
	so we get a $\Omega(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	\end{proof}

@@

\begin{theorem}[MAP Bound]
Consider an exponential distribution with sufficient statistics coming from a gamma distribution $\Gamma(\alpha, \beta^*)$ with mean parameter $\mu^* = \frac{\alpha}{\beta^*}$.
The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP  with prior hyper-parameters $(n_0,\mu_0)$ is upper bounded as
\alignn{
    \forall n\geq \inv{\alpha},\quad \quad
	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
	\leq\frac{1}{n_0+n} + 
	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} - 1 }^2 }{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} +  \frac{n - \inv{\alpha}}{n_0}}}
}
\end{theorem} 
Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
Expressions for $n\alpha < 1$ are less elegant as shown below:
\alignn{
 	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
	\leq \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}} -1 + \frac{a - n_0}{n_0+n}
	\quadtext{where}
	 a:=n_0\frac{\mu_0}{\mu^*} \; .
}
